Phylodynamic inference of contact network parameters through approximate Bayesian computation

### 2016-06-14 doc/thesis/methods_results.Rnw
Started writing intro for BA methods section

\section{\software{Netabc}: a computer program for estimation of contact
network parameters with kernel-assisted ABC}
\label{sec:netabc}

\software{Netabc} is a computer program to perform statistical inference of
contact network parameters from an estimated transmission tree using
\gls{ABC}. {\color{blue}\uline{As discussed in }\cref{sec:obj}, \uline{the principal
statistical algorithm used by \software{netabc} is adaptive
\gls{ABC}-\gls{SMC}~\autocite{del2012adaptive}. In addition, there are two
supplementary components which are specific to the domain of phylogenetics and
contact networks: Gillespie simulation~\autocite{gillespie1976general}, to
simulate transmission trees on contact networks; and the tree
kernel~\autocite{poon2013mapping}, which is used as the distance function in
\gls{ABC} to compare transmission trees~\autocite{poon2015phylodynamic}} (see
\cref{sec:abc}).} We give a high-level overview of the program here, before
describing these components in detail. \software{Netabc} takes as input an
estimated transmission tree, which can be derived from a viral phylogeny by
rooting and time-scaling as described in \cref{subsec:phylodynamics} or
estimated by other methods~\autocite{cottam2008integrating,
jombart2011reconstructing, ypma2012unravelling, morelli2012bayesian,
didelot2014bayesian, hall2015epidemic}. We variously refer to this estimated
transmission tree as the observed tree, input tree, or true tree.

As described in \cref{sec:smc}, \software{netabc} keeps track of a population
of particles $x^{(k)}$, each of which contains particular parameter values
$\theta^{(k)}$ for the {\color{blue}\uline{contact network}} model we are
trying to fit {\color{blue}\uline{to the input tree}}. A small number of
contact networks $z^{(k)}$ are generated under the model for each particle, in
accordance with that particle's parameters. An epidemic is simulated over each
of these networks using Gillespie simulation, and by keeping track of its
progress, a transmission tree is obtained. Thus, each particle becomes
associated with several simulated transmission trees. These trees are compared
to the input tree using the tree kernel. Particles are weighted according to
the similarity of their associated simulated trees with the true tree, with
more similar trees receiving higher weights. The particles are iteratively
perturbed to explore the parameter space, and particles with simulated trees
too distant from the true tree are periodically dropped and resampled. Once a
convergence criterion is attained, the final set of particles is used as a
Monte Carlo approximation to the target distribution of \gls{ABC}, which is
assumed to resemble the posterior distribution on model parameters (see
\cref{sec:abc}). A graphical schematic of this algorithm is given in
\cref{fig:abcsmc}.

\begin{figure}
    \includegraphics{abc-smc.pdf}
    \caption[Graphical schematic of the ABC-SMC algorithm implemented in \software{netabc}.]{
      Graphical schematic of the \gls{ABC}-\gls{SMC} algorithm implemented in
      \software{netabc}. Particles are initially drawn from their prior
      distributions, making the initial population a Monte Carlo approximation
      to the prior. At each iteration, particles are perturbed, and a distance
      threshold around the true tree contracts. Particles are rejected, and
      eventually resampled, when all their associated simulated trees lie
      outside the threshold. As the algorithm progresses, the population
      smoothly approaches a Monte Carlo approximation of the \gls{ABC} target
      distribution, which is assumed to resemble the posterior.
    }
    \label{fig:abcsmc}
\end{figure}

\software{Netabc} is written in the \software{C} programming language. The
\software{igraph} library~\autocite{csardi2006igraph} is used to generate and
store contact networks and phylogenies. Judy arrays~\autocite{baskins2004judy}
are used for hash tables and dynamic programming matrices. The
\gls{GSL}~\autocite{gough2009gnu} is used to generate random draws from
probability distributions, and to perform the bisection step in the adaptive
\gls{ABC}-\gls{SMC} algorithm. Parallelization is implemented with \gls{POSIX}
threads~\autocite{barney2009posix}. In addition to the \software{netabc} binary
to perform kernel-assisted \gls{ABC}, we provide three additional stand-alone
utilities: \software{treekernel}, to calculate the tree kernel;
\software{nettree}, to simulate a transmission tree over a contact network; and
\software{treestat}, to compute various summary statistics of phylogenies. The
programs are freely available at \url{https://github.com/rmcclosk/netabc}.

To check that our implementation of Gillespie simulation was correct, we
reproduced Figure 1A of \textcite{leventhal2012inferring} (our
\cref{fig:leventhal}), which plots the imbalance of transmission trees
simulated over four network models at various levels of pathogen
transmissibility. Our implementation of adaptive \gls{ABC}-\gls{SMC} was tested
by applying it to the same mixture of Gaussians used by
\textcite{del2012adaptive} to demonstrate their method (originally used
by~\textcite{sisson2007sequential}). We were able to obtain a close
approximation to the function (see \cref{fig:smctest}), and attained the
stopping condition used by the authors in a comparable number of steps. To
check that the algorithm would converge to a bimodal distribution, we also
applied it to a mixture of two Gaussians with means $\pm$4 and variances 1. The
algorithm was able to recover both peaks (\cref{fig:smctest2}).

@@

The simulation of epidemics, and the corresponding transmission trees, over
contact networks is performed in \software{netabc} using the Gillespie
simulation algorithm~\autocite{gillespie1976general}. This method has been
independently implemented and applied by several
authors~\autocite[\textit{e.g.}][]{o2011contact, robinson2013dynamics,
leventhal2012inferring, groendyke2011bayesian, villandre2016assessment}.
\textcite{groendyke2011bayesian} published their implementation as an
\software{R} package, but since the \gls{SMC} algorithm is quite
computationally intensive, we chose to implement our own version in
\software{C} {\color{blue}\uline{as part of \software{netabc}}}.

Let $G = (V, E)$ be a directed contact network. We assume the individual nodes
and edges of $G$ follow the dynamics of the \gls{SIR}
model~\autocite{kermack1927contribution}. Each directed edge $e = (u, v)$ in
the network is associated with a transmission rate $\beta_e$, which indicates
that, once $u$ becomes infected, the waiting time until $u$ infects $v$ is
distributed as $\Exponential(\beta_e)$. Note that $v$ may become infected
before this time has elapsed, if $v$ has other incoming edges. $v$ also has a
removal rate $\gamma_v$, so that the waiting time until removal of $v$ from the
population is $\Exponential(\gamma_v)$. Removal may correspond to death or
recovery with immunity, or a combination of both, but in our implementation
recovered nodes never re-enter the susceptible population. We define a
\defn{discordant edge} as an edge $(u, v)$ where $u$ is infected and $v$ has
never been infected.

To describe the algorithm, we introduce some notation and variables. Let
$\inc(v)$ be the set of incoming edges to $v$, and $\out(v)$ be the set of
outgoing edges from $v$. Let $I$ be the set of infected nodes in the network,
$R$ be the set of removed nodes, and $S$ be the remaining susceptible nodes,
and $D$ be the set of discordant edges in the network. Let $\beta$ be the total
transmission rate over all discordant edges, and $\gamma$ be the total removal
rate of all infected nodes,
\[
  \beta = \sum_{e \in D} \beta_e, \quad
  \gamma = \sum_{v \in I} \gamma_v.
\]
The variables $S$, $I$, $R$, $D$, $\beta$, and $\gamma$ are all updated as the
simulation progresses. When a node $v$ becomes infected, it is deleted from $S$
and added to $I$. Any formerly discordant edges in $\inc(v)$ are deleted from
$D$, and edges in $\out(v)$ to nodes in $S$ are added to $D$. If $v$ is later
removed, it is deleted from $I$ and added to $R$, and any discordant edges in
$\out(v)$ are deleted from $D$. At the time of either infection or removal, the
variables $\beta$ and $\gamma$ are updated to reflect the changes in the
network. Since these updates are straightforward, we do not write them
explicitly in the algorithm.

@@

The Gillespie simulation algorithm is given as Algorithm~\ref{alg:nettree}. The
transmission tree $T$ is simulated along with the epidemic. We keep a map
called $\tip$, which maps infected nodes in $I$ to the tips of $T$. The
simulation continues until either there are no discordant edges left in the
network, or we reach a user-defined cutoff of time ($t_{\max}$) or number of
infections ($I_{\max}$). We use the notation $\Uniform(0, 1)$ to indicate a
number drawn from a uniform distribution on $(0, 1)$, and likewise for
$\Exponential(\lambda)$. The combined number of internal nodes and tips in $T$
is denoted $|T|$.

\begin{algorithm}
  \label{alg:nettree}
  \caption{Simulation of an epidemic and transmission tree over a contact network}
  \begin{algorithmic}
    \State infect a node $v$ at random, updating $S$, $I$, $D$, $\beta$ and $\gamma$
    \State $T \gets$ a single node with label $1$
    \State $\tip[v] \gets 1$
    \State $t \gets 0$
    \While{$D \neq \emptyset$ and $|I| + |R| < I_{\max}$ and $t < t_{\max}$}
      \State $s \gets \min(t_{\max} - t, \Exponential(\beta + \gamma))$
      \For{$v \in \tip$}
        \State{extend the branch length of $\tip[v]$ by $s$}
      \EndFor
      \State $t \gets t + s$
      \If{$t < t_{\max}$}
        \If{$\Uniform(0, \beta + \gamma) < \beta$}
          \State choose an edge $e = (u, v)$ from $D$ with probability $\beta_e / \beta$
                 and infect $v$
          \State $\tip[v] \gets |T|+1$
          \Comment{add new tips to tree and tip array}
          \State $\tip[u] \gets |T|+2$
          \Comment{corresponding to $u$ and $v$}
          \State add tips with labels $(|T|+1)$ and $(|T|+2)$ to $T$
          \State connect the new nodes to $\tip[v]$ in $T$, with branch lengths $0$
        \Else
          \State choose a node $v$ from $I$ with probability $\gamma_v / \gamma$
                 and remove $v$
          \State delete $v$ from $\tip$
        \EndIf
        \State update $S$, $I$, $R$, $D$, $\beta$, and $\gamma$
      \EndIf
    \EndWhile
  \end{algorithmic}
\end{algorithm}

@@

The tree kernel developed by \textcite{poon2013mapping} provides a
comprehensive similarity score between two phylogenetic trees, via the
dot-product of the two trees' feature vectors in the space of all possible
subset trees with branch lengths (see \cref{subsec:treeshape}). {\color{blue}
\uline{Because the branch lengths are continuous, there are infinitely many
possible subset trees; hence, the feature space is infinite-dimensional.}}
The kernel was implemented using the fast algorithm developed by
\textcite{moschitti2006making}. First, the production rule of each node, which
is the total number of children and the number of leaf children, is recorded.
The nodes of both trees are ordered by production rule, and a list of pairs of
nodes sharing the same production rule is created. These are the nodes for
which the value of the tree kernel must be computed - all other pairs have a
value of zero. The pairs to be compared are then re-ordered so that the child
nodes are always evaluated before their parents. Due to its recursive
definition, ordering the pairs in this way allows the tree kernel to be
computed by dynamic programming. The complexity of this implementation is
$O(|T_1||T_2|)$ for the two trees $T_1$ and $T_2$ being compared.

The tree kernel cannot be used directly as a distance measure for \gls{ABC},
since it is maximized, not minimized, when the two trees being compared are the
same. Therefore, we defined the distance between two trees as
\[
  \rho(T_1, T_2) = 1 - \frac{K(T_1, T_2)}{\sqrt{K(T_1, T_1) K(T_2, T_2)}},
\]
which is a number between 0 and 1 that is minimized when $T_1 = T_2$. This is
similar to the normalization used by \textcite{collins2002new,
poon2013mapping}.

@@

We implemented the adaptive \gls{SMC} algorithm for \gls{ABC} developed by
\textcite{del2012adaptive}. This algorithm is similar to the reference
\gls{ABC}-\gls{SMC} algorithm described in \cref{subsec:abcalg}, except that
the sequence of tolerances $\varepsilon_i$ is automatically determined rather
than specified in advance. The tolerances are chosen such that the \gls{ESS} of
the particle population, which indicates the quality of the Monte Carlo
approximation (see \cref{subsec:sis}), decays at a controlled rate. A
sudden precipitous drop in \gls{ESS} would indicate that only a small number of
particles had non-zero importance weights, which would result in a very poor
Monte Carlo approximation to the target distribution. This situation is
referred to as the ``collapse'' of the approximation, {\color{blue}\uline{also
known as particle degeneracy} (see \cref{subsec:smc})} and is mitigated by the
adaptive approach. A single parameter {\color{red}\sout{$\alpha$ (not to be
confused with the \gls{BA} model parameter)}} controls the decay rate.
{\color{blue}\uline{In the original paper of }\textcite{del2012adaptive},
\uline{the parameter is called $\alpha$, but to avoid confusion with the
\gls{BA} parameter of the same name, we will refer to it here as
\gls{alphaess}.}}
The tolerance $\varepsilon_i$ is chosen to satisfy
\[
  \ESS(w_i) = \alpha_{\ESS} \ESS(w_{i-1}),
\]
where, $w_i$ is the vector of weights at the $i$th step. Note that, since $w_i$
depends on $\varepsilon_i$, this equation solves for the updated weights and
the updated tolerance simultaneously. As pointed out by
\textcite{del2012adaptive}, the equation has no analytic solution, but can be
solved numerically by bisection. The forward kernels $K_i$ are taken to be
\gls{MCMC} kernels with stationary distributions $\pi_{\varepsilon_i}$ and
proposal distributions
\[
  q_i(\theta, \theta') \prod_{k=1}^M \Pr(z_i^{(k)'} \mid \theta'),
\]
where $\theta$ is the vector of model parameters and $z_k$ are $M$ datasets
simulated according to $\theta'$. In our implementation, $q$ is either a
Gaussian proposal for continuous parameters, or a Poisson proposal for discrete
parameters. For the Poisson proposals, the number of steps to move the particle
is drawn from a Poisson distribution, and the direction in which to move the
particle is chosen uniformly at random. For both proposals, the variance was
set equal to twice the empirical variance of the particles,
following~\autocite{beaumont2009adaptive, del2012adaptive}. The backwards
kernels are
\[
  L_{i-1}(x', x) = \frac{\pi_n(x)K(x, x')}{\pi_n(x')}.
\]
When substituted into \cref{eq:smcwt}, the forward kernels $K(x, x')$ and
densities $\pi_n(x') = \pi_{\varepsilon_n}(x')$ cancel out, and we are left
with the weight update 
\begin{align*}
  w_i(x) 
    &\propto w_{i-1}(x) \frac{\pi_n(x \mid y)}{\pi_{i-1}(x \mid y)} \\
    &= w_{i-1}(x) \frac{\pi(x) \pi_i(y \mid x)}{\pi(x) \pi_{i-1}(y \mid x)} \\
    &= w_{i-1}(x) \frac{\sum_{k=i}^M \I_{A_{\varepsilon_i, y}}(z_k)}
            {\sum_{k=i}^M \I_{A_{\varepsilon_{i-1}, y}}(z_k)}.
\end{align*}
In other words, when the distance threshold $\varepsilon_{i-1}$ is contracted
to $\varepsilon_i$, the particles' weights are multiplied by the proportion of
simulated datasets that are still inside the new threshold. 
The user may specify a final tolerance $\varepsilon$, or a final acceptance
rate of the \gls{MCMC} kernel, and the algorithm will be stopped when either of
these termination conditions is reached. The latter condition stops the
algorithm when the particles are not moving around very much, implying little
change in the estimated target.

@@

{\color{blue}\uline{We present here a non-rigorous justification for the use of
\gls{ABC} for the problem at hand, as opposed to more frequently-used
approaches for fitting mathematical models} (see \cref{chp:prelim}).
\uline{Consider a contact network model with parameters $\theta$, and an
estimated transmission tree $T$. Taking a Bayesian approach, our aim is to
obtain a sample from the posterior distribution on the model's parameters given
our data,}
\[
    \Pr(\theta \mid T) = \frac{\Pr(T \mid \theta) \Pr(\theta)}{\Pr(T)}.
\]
\uline{For all but the simplest models, the normalizing constant $\Pr(T)$ is an
intractable integral. What we shall argue here is that, in contrast to most
commonly studied mathematical models, the likelihood $\Pr(T \mid \theta)$ is
also likely to be intractible in our case.}

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{transtree}
    \caption[
        Illustration of an estimated transmission tree without labels and two
        possible underlying complete transmission trees with labels.
    ]{
        Illustration of an estimated transmission tree without labels (left)
        and two possible underlying complete transmission trees with labels
        (right). In the top right scenario, the epidemic began with node $a$
        who transmitted first to $b$ and then to $c$. In the bottom right
        scenario, $b$ was the index case; $b$ infected $c$, who went on to
        infect $a$. A transmission tree estimated from a viral phylogeny would
        have the same topology and tip labels in both cases.
    }
    \label{fig:tt}
\end{figure}

>>>>

\section{\software{Netabc}: a computer program for estimation of contact
network parameters with ABC}
\label{sec:netabc}

\software{Netabc} is a computer program to perform statistical inference of
contact network parameters from an estimated transmission tree using
\gls{ABC}. \add{As discussed in \cref{sec:obj}, the principal statistical
algorithm used by \software{netabc} is adaptive
\gls{ABC}-\gls{SMC}~\autocite{del2012adaptive}. In addition, there are two
supplementary components which are specific to the domain of phylogenetics and
contact networks: Gillespie simulation~\autocite{gillespie1976general}, to
simulate transmission trees on contact networks; and the tree
kernel~\autocite{poon2013mapping}, which is used as the distance function in
\gls{ABC} to compare transmission trees~\autocite{poon2015phylodynamic} (see
\cref{sec:abc}).} We give a high-level overview of the program here, before
describing these components in detail. \software{Netabc} takes as input an
estimated transmission tree, which can be derived from a viral phylogeny by
rooting and time-scaling as described in \cref{subsec:phylodynamics} or
estimated by other methods~\autocite{cottam2008integrating,
jombart2011reconstructing, ypma2012unravelling, morelli2012bayesian,
didelot2014bayesian, hall2015epidemic}. We variously refer to this estimated
transmission tree as the observed tree, input tree, or true tree.

As described in \cref{sec:smc}, \software{netabc} keeps track of a population
of particles $x^{(k)}$ indexed by an integere $k$, each of which contains
particular parameter values $\theta^{(k)}$ for the \add{contact network} model
we are trying to fit \add{to the input tree}. A small number \gls{M} of contact
networks $z^{(k,i)}$, $1 \leq i \leq \gls{M}$, are generated under the model
for each particle in accordance with that particle's parameters. An epidemic is
simulated over each of these networks using Gillespie simulation, and by
keeping track of its progress, a transmission tree is obtained. Thus, each
particle becomes associated with several simulated transmission trees. These
trees are compared to the input tree using the tree kernel. Particles are
weighted according to the similarity of their associated simulated trees with
the true tree, with more similar trees receiving higher weights. The particles
are iteratively perturbed to explore the parameter space, and particles with
simulated trees too distant from the true tree are periodically dropped and
resampled. Once a convergence criterion is attained, the final set of particles
is used as a Monte Carlo approximation to the target distribution of \gls{ABC},
which is assumed to resemble the posterior distribution on model parameters
(see \cref{sec:abc}). A graphical schematic of this algorithm is shown in
\cref{fig:abcsmc}.

\begin{figure}
    \includegraphics{abc-smc.pdf}
    \caption[
        Graphical schematic of the ABC-SMC algorithm implemented in \software{netabc}.
    ]{
        Graphical schematic of the \gls{ABC}-\gls{SMC} algorithm implemented in
        \software{netabc}. Particles are initially drawn from their prior
        distributions, making the initial population a Monte Carlo
        approximation to the prior. At each iteration, particles are perturbed,
        and a distance threshold around the true tree contracts. Particles are
        rejected, and eventually resampled, when all their associated simulated
        trees lie outside the threshold. As the algorithm progresses, the
        population smoothly approaches a Monte Carlo approximation of the
        \gls{ABC} target distribution, which is assumed to resemble the
        posterior.
    }
    \label{fig:abcsmc}
\end{figure}

\software{Netabc} is written in the \software{C} programming language. The
\software{igraph} library~\autocite{csardi2006igraph} is used to generate and
store contact networks and phylogenies. Judy arrays~\autocite{baskins2004judy}
are used for hash tables and dynamic programming matrices. The
\gls{GSL}~\autocite{gough2009gnu} is used to generate random draws from
probability distributions, and to \del{perform the bisection step} \add{solve
for the next $\varepsilon$ by bisection} in the adaptive \gls{ABC}-\gls{SMC}
algorithm. Parallelization is implemented with \gls{POSIX}
threads~\autocite{barney2009posix}. In addition to the \software{netabc} binary
to perform \gls{ABC}, we provide three additional stand-alone utilities:
\software{treekernel}, to calculate the tree kernel; \software{nettree}, to
simulate a transmission tree over a contact network; and \software{treestat},
to compute various summary statistics of phylogenies. The programs are freely
available at \url{https://github.com/rmcclosk/netabc}.

To check that our implementation of Gillespie simulation was correct, we
reproduced Figure 1A of \textcite{leventhal2012inferring} (our
\cref{fig:leventhal}), which plots the imbalance of transmission trees
simulated over four network models at various levels of pathogen
transmissibility. Our implementation of adaptive \gls{ABC}-\gls{SMC} was tested
by applying it to the same mixture of Gaussians used by
\textcite{del2012adaptive} to demonstrate their method (originally used
by~\textcite{sisson2007sequential}). We were able to obtain a close
approximation to the function (see \cref{fig:smctest}), and attained the
stopping condition used by the authors in a comparable number of steps. To
check that the algorithm would converge to a bimodal distribution, we also
applied it to a mixture of two Gaussians with means $\pm4$ and variances 1. The
algorithm was able to recover both peaks (\cref{fig:smctest2}).

@@

The simulation of epidemics, and the corresponding transmission trees, over
contact networks is performed in \software{netabc} using the Gillespie
simulation algorithm~\autocite{gillespie1976general}. This method has been
independently implemented and applied by several
authors~\autocite[\textit{e.g.}][]{o2011contact, robinson2013dynamics,
leventhal2012inferring, groendyke2011bayesian, villandre2016assessment}.
\textcite{groendyke2011bayesian} published their implementation as an
\software{R} package, but since the \gls{SMC} algorithm is quite
computationally intensive, we chose to implement our own version in
\software{C} \add{as part of \software{netabc}}.

Let $G = (V, E)$ be a directed contact network. We assume the individual nodes
and edges of $G$ follow the dynamics of the \gls{SIR}
model~\autocite{kermack1927contribution}. Each directed edge $e = (u, v)$ in
the network is associated with a transmission rate $\beta_e$, which indicates
that, once $u$ becomes infected, the waiting time until $u$ infects $v$ is
distributed as $\Exponential(\gls{beta}_e)$. Note that $v$ may become infected
before this time has elapsed, if $v$ has other incoming edges. $v$ also has a
removal rate $\gls{nu}_v$, so that the waiting time until removal of $v$ from
the population is $\Exponential(\gls{nu}_v)$. Removal may correspond to death
or recovery with immunity, or a combination of both, but in our implementation
recovered nodes never re-enter the susceptible population. We define a
\defn{discordant edge} as an edge $(u, v)$ where $u$ is infected and $v$ has
never been infected. \add{In the epidemiology literature, the symbol $\gamma$
is usually used in place of \gls{nu}; we use \gls{nu} here to distinguish the
recovery rate from the power law exponent of scale free networks (see
\cref{subsec:pa}).}

To describe the algorithm, we introduce some notation and variables. Let
$\inc(v)$ be the set of incoming edges to $v$, and $\out(v)$ be the set of
outgoing edges from $v$. Let \gls{gI} be the set of infected nodes in the
network, \gls{gR} be the set of removed nodes, \gls{gS} be the set of
susceptible nodes, and \gls{gD} be the set of discordant edges in the network.
Let \gls{beta} be the total transmission rate over all discordant edges, and
\gls{nu} be the total removal rate of all infected nodes,
\[
    \gls{beta} = \sum_{e \in \gls{gD}} \gls{beta}_e, \quad
    \gls{nu} = \sum_{v \in \gls{gI}} \gls{nu}_v.
\]
The variables \gls{gS}, \gls{gI}, \gls{gR}, \gls{gD}, \gls{beta}, and \gls{nu}
are all updated as the simulation progresses. When a node $v$ becomes infected,
it is deleted from \gls{gS} and added to \gls{gI}. Any formerly discordant
edges in $\inc(v)$ are deleted from \gls{gD}, and edges in $\out(v)$ to nodes
in \gls{gS} are added to \gls{gD}. If $v$ is later removed, it is deleted from
\gls{gI} and added to \gls{gR}, and any discordant edges in $\out(v)$ are
deleted from \gls{gD}. At the time of either infection or removal, the
variables \gls{beta} and \gls{nu} are updated to reflect the changes in the
network. \del{The updates to \gls{gS}, \gls{gI}, \gls{gR}, \gls{gD},
\gls{beta}, and \gls{nu} are straightforward and are not written explicitly in
the algorithm.}

@@

The Gillespie simulation algorithm is given as \cref{alg:nettree}. The
transmission tree \gls{T} is simulated along with the epidemic. We keep a map
called ``$\tip$'', which maps infected nodes in \gls{gI} to the tips of
\gls{T}. The simulation continues until either there are no discordant edges
left in the network, or we reach a user-defined cutoff of time (\gls{tmax}) or
number of infections (\gls{I}). We use the notation $\Uniform(0, 1)$ to
indicate a number drawn from a uniform distribution on $(0, 1)$, and
$\Exponential(\lambda)$ to indicate a number drawn from an exponential
distribution with rate $\lambda$. The combined number of internal nodes and
tips in \gls{T} is denoted $|T|$. \add{The updates to \gls{gS}, \gls{gI},
\gls{gR}, \gls{gD}, \gls{beta}, and \gls{nu} described in the previous
paragraph are not written explicitly in \cref{alg:nettree}, as they are quite
straightforward and would only obfuscate the pseudocode.}

\begin{algorithm}
    \label{alg:nettree}
    \caption{Simulation of an epidemic and transmission tree over a contact network}
    \begin{algorithmic}
        \State infect a node $v$ at random, updating \gls{gS}, \gls{gI},
                 \gls{gD}, \gls{beta}, and \gls{gamma}
        \State $\gls{T} \gets$ a single node with label $1$
        \State $\tip[v] \gets 1$
        \State $\gls{t} \gets 0$
        \While{$\gls{gD} \neq \emptyset$ and $|\gls{gI}| + |\gls{gR}| < \gls{I}$
                and $\gls{t} < \gls{tmax}$}
            \State $s \gets \min(\gls{tmax} - \gls{t},\,
                \Exponential(\gls{beta} + \gls{nu}))$
            \For{$v \in \tip$}
                \State{extend the branch length of $\tip[v]$ by $s$}
            \EndFor
            \State $\gls{t} \gets \gls{t} + s$
            \If{$\gls{t} < \gls{tmax}$}
                \If{$\Uniform(0, \gls{beta} + \gls{nu}) < \gls{beta}$}
                    \State choose an edge $e = (u, v)$ from \gls{gD} with
                        probability $\gls{beta}_e / \gls{beta}$ and infect $v$
                    \State $\tip[v] \gets |\gls{T}|+1$
                    \Comment add new tips to tree and tip array
                    \State $\tip[u] \gets |\gls{T}|+2$
                    \Comment corresponding to $u$ and $v$ 
                    \State add tips with labels $(|\gls{T}|+1)$ and
                        $(|\gls{T}|+2)$ to \gls{T}
                    \State connect the new nodes to $\tip[v]$ in $\gls{T}$,
                        with branch lengths $0$ 
                \Else
                    \State choose a node $v$ from \gls{gI} with probability
                        $\gls{nu}_v / \gls{nu}$ and remove $v$
                    \State delete $v$ from $\tip$
                \EndIf
                \State update \gls{gS}, \gls{gI}, \gls{gR}, \gls{gD},
                    \gls{beta}, and \gls{nu}
            \EndIf
        \EndWhile
    \end{algorithmic}
\end{algorithm}

@@

The tree kernel developed by \textcite{poon2013mapping} provides a
comprehensive similarity score between two phylogenetic trees, via the
dot-product of the two trees' feature vectors in the space of all possible
subset trees with branch lengths (see \cref{subsec:treeshape}). \add{Because
the branch lengths are continuous, there are infinitely many possible subset
trees; hence, the feature space is infinite-dimensional.} The kernel was
implemented using the fast algorithm developed by
\textcite{moschitti2006making}. First, the production rule of each node, which
is the total number of children and the number of leaf children, is recorded.
The nodes of both trees are ordered by production rule, and a list of pairs of
nodes sharing the same production rule is created. These are the nodes for
which the value of the tree kernel must be computed - all other pairs have a
value of zero. The pairs to be compared are then re-ordered so that the child
nodes are always evaluated before their parents. Due to its recursive
definition, ordering the pairs in this way allows the tree kernel to be
computed by dynamic programming. The complexity of this implementation is
$O(|\gls{T}_1||\gls{T}_2|)$ for the two trees $\gls{T}_1$ and $\gls{T}_2$ being
compared.

The tree kernel cannot be used directly as a distance measure for \gls{ABC},
since it is maximized, not minimized, when the two trees being compared are the
same. Therefore, we defined the distance between two trees as
\[
    \gls{rho}(\gls{T}_1, \gls{T}_2) = 1 - \frac{K(\gls{T}_1, \gls{T}_2)}{\sqrt{K(\gls{T}_1, \gls{T}_1) K(\gls{T}_2, \gls{T}_2)}},
\]
which is a number between 0 and 1 that is minimized when $\gls{T}_1 = \gls{T}_2$. This is
similar to the normalization used by \textcite{collins2002new,
poon2013mapping}.

@@

We implemented the adaptive \gls{SMC} algorithm for \gls{ABC} developed by
\textcite{del2012adaptive}. This algorithm is similar to the reference
\gls{ABC}-\gls{SMC} algorithm described in \cref{subsec:abcalg}, except that
the sequence of tolerances $\gls{epsilon}_i$ is automatically determined rather
than specified in advance. The tolerances are chosen such that the \gls{ESS} of
the particle population, which indicates the quality of the Monte Carlo
approximation (see \cref{subsec:sis}), decays at a controlled rate. A
sudden precipitous drop in \gls{ESS} would indicate that only a small number of
particles had non-zero importance weights, which would result in a very poor
Monte Carlo approximation to the target distribution. This situation is
referred to as the collapse of the approximation \add{or particle degeneracy
(see \cref{subsec:smc})} and is mitigated by the adaptive approach. A single
parameter \del{$\alpha$ (not to be confused with the \gls{BA} model parameter)}
controls the decay rate. \add{In the original paper of
\textcite{del2012adaptive}, the parameter is called $\alpha$, but to avoid
confusion with the \gls{BA} parameter of the same name we will refer to it
here as \gls{alphaess}.} The tolerance $\gls{epsilon}_i$ is chosen to satisfy
\[
    \ESS(w_i) = \gls{alphaess} \ESS(w_{i-1}),
\]
where, $w_i$ is the vector of weights at the $i$th step. Note that, since $w_i$
depends on $\gls{epsilon}_i$, this equation solves for the updated weights and
the updated tolerance simultaneously. As pointed out by
\textcite{del2012adaptive}, the equation has no analytic solution, but can be
solved numerically by bisection. The forward kernels $K_i$ are taken to be
\gls{MCMC} kernels with stationary distributions $\pi_{\gls{epsilon}_i}$ and
proposal distributions
\[
    \gls{q}_i\left(\theta^{(k)}, \theta^{(k)\prime}\right) 
    \prod_{j=1}^{\gls{M}} \lik\left(z^{(j,k)\prime} \middle| \theta^{(k)\prime} \right),
\]
where $\theta^{(k)}$ is the vector of model parameters \add{associated with
particle $x^{(k)}$} and $z^{(j,k)\prime}$, $1 \leq j \leq \gls{M}$, are \gls{M}
datasets simulated according to $\theta^{(k)\prime}$. In our implementation,
the $\gls{q}_i$ \add{are constructed component-wise for $\theta$ out of}
Gaussian proposals for continuous parameters and Poisson proposals for discrete
parameters. For the Poisson proposals, the number of \add{discrete} steps to
move the particle is drawn from a Poisson distribution, and the direction in
which to move the particle is chosen uniformly at random. The variance of each
proposal distribution was set equal to twice the empirical variance of the
particles, following~\autocite{beaumont2009adaptive, del2012adaptive}. The
backwards kernels are
\[
    L_{i-1}(x', x) = \frac{\pi_i(x)K_i(x, x')}{\pi_i(x')}.
\]
Here we have written $x'$ for $x_i$ and $x$ for $x_{i-1}$ to emphasize that
$x_{i-1}$ is the current value of the particle and $x_i$ is the proposed value.
When substituted into \cref{eq:smcwt}, the forward kernels $K_i(x, x')$ and
densities $\pi_i(x') = \pi_{\varepsilon_i}(x')$ cancel out, and we are left
with the following weight update.
\begin{align*}
    w_i(x) 
    & \madd{\propto w_{i-1}(x) \frac{\pi_i(x') L_{i-1}(x', x)}{\pi_{i-1}(x)K_i(x, x')}}
        & \madd{\text{importance weight update from SMC}} \\
    &\madd{=w_{i-1}(x) \frac{\pi_i(x') \pi_i(x) K_i(x, x')}{\pi_{i-1}(x)K_i(x, x')\pi_i(x')}}
        & \madd{\text{substitute } L_{i-1}} \\
    &= w_{i-1}(x) \frac{\pi_i(x)}{\pi_{i-1}(x)}
        & \madd{\text{cancel } K_i(x, x') \text{ and } \pi_i(x')} \\
    &= \madd{w_{i-1}(x) \frac{\pi(\theta) \prod_{j=1}^M f(z^{(j)\prime} \mid \theta) 
                        \sum_{j=i}^M \I_{A_{\varepsilon_i, y}}(z^{(j)})}
                       {\pi(\theta) \prod_{j=1}^M f(z^{(j)\prime} \mid \theta) 
                       \sum_{j=i}^M \I_{A_{\varepsilon_{i-1}, y}}(z^{(j)})}}
        & \madd{\text{definition of } \pi_i(x)} \\
    &= w_{i-1}(x) \frac{\sum_{j=i}^M \I_{A_{\varepsilon_i, y}}(z^{(j)})}
        {\sum_{j=i}^M \I_{A_{\varepsilon_{i-1}, y}}(z^{(j)})}
        & \madd{\text{cancel prior and likelihood.}}
\end{align*}
In other words, when the distance threshold $\gls{epsilon}_{i-1}$ is contracted
to $\gls{epsilon}_i$, the particles' weights are multiplied by the proportion of
simulated datasets that are still inside the new threshold. The user may
specify a final tolerance $\gls{epsilon}$, or a final acceptance rate of the
\gls{MCMC} kernel, and the algorithm will be stopped when either of these
termination conditions is reached. The latter condition stops the algorithm
when the particles are not moving around very much, implying little change in
the estimated target.

@@

\add{We present here a non-rigorous justification for the use of
\gls{ABC} for the problem at hand, as opposed to more frequently-used
approaches for fitting mathematical models (see \cref{chp:prelim}). Consider a
contact network model with parameters $\theta$, and an estimated transmission
tree \gls{T}. Taking a Bayesian approach, our aim is to obtain a sample from
the posterior distribution on the model's parameters given our data,}
\[
    \madd{\post(\theta \mid \gls{T}) = \frac{\lik(\gls{T} \mid \theta) \prior(\theta)}
    {\int_{\Theta}\lik(\gls{T} \mid \theta) \prior(\theta) \d\theta}.}
\]
\add{For all but the simplest models, the normalizing constant in the
denominator is an intractable integral. What we shall argue here is that, in
contrast to most commonly studied mathematical models, the likelihood
$\lik(\gls{T} \mid \theta)$ is also likely to be intractible in our case.}
